---
title: "Medical Insurance Charges Analysis"
subtitle: "Understanding Demographics and Smoking behaviours with Statistical Analysis"
author: "Jialun Darren Huang"
output:
  slidy_presentation: 
    font_adjustment: 0
    highlight: haddock
    mathjax: "default"
    df_print: paged
    includes:
      in_header: header.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	fig.align = "center",
	message = FALSE,
	warning = FALSE
)

#
library(dplyr) # For data manipulation
library(tidyr) # For reshaping data
library(car) # For checking assumptions in two sampled t-test
library(ggplot2) # For creating plots
library(patchwork) # For combining plots side by side
library(knitr) # Useful for creating nice tables
library(kableExtra) # Editing nice kable tables
```

# Introduction

- Medical insurance helps individuals manage financial burden of healthcare expenses
- Insurance charges can differ with individual demographics and lifestyle characteristics.
- Understanding these differences can help us gain insights to:
  - What are some factors that associate to higher medical insurance costs
  - Insurers and policymakers: Design a fairer, unbiased, and more data driven insurance pricing model
  - Individuals with medical insurance: How their own lifestyle can impact medical and their insurance costs, so they can choose better lifestyle choices
  
# Problem Statement

- This presentation investigates whether smoking behaviour is associated with demographic factors and higher medical insurance charges

- Specifically, we look into:
  - Whether Sex and region are associated with smoking status
  - Whether smokers have a higher medical insurance charges, compared to non-smokers

- These relationships would be analysed using
  - Categorical association with Chi-square tests
  - Hypothesis testing, comparing differences in medical insurance charges between groups
  
# Data

```{r include = FALSE}
ori_df <- read.csv("insurance.csv")
df <- ori_df[, c("smoker", "sex", "region", "charges")]
```

```{r include = FALSE}
df$smoker <- factor(df$smoker, 
                    levels = c("no", "yes"))
df$sex <- factor(df$sex, 
                 levels = c("female", "male"))
df$region <- factor(df$region, 
                 levels = c("southwest", "southeast", "northwest", "northeast"))

smoker_color_code <- c("no" = "lightgreen", "yes" = "orange2")
sex_color_code <- c("female" = "lightpink", "male" = "cornflowerblue")
region_color_code <- c("southwest" = "lightpink", "southeast" = "cornflowerblue",
                       "northwest" = "lightgreen", "northeast" = "orange2")
```

- Medical Insurance Cost Dataset from Kaggle by user Mosap Abdel-Ghany (See references). 
- Dataset contains individual medical insurance records with demographic and lifestyle information
- No explicit sampling method was said in the data source, however, it could have been sampled (e.g. from a registry or larger population), as the data contains just over 1000 insurance records.
- Subset of features was selected for the analysis: sex, region, smoker, charges. (See Table 1 for Dataset Description)
- Conducted preprocessing steps to ensure quality of data and its integrity
  - Checked for any missing values in the features
  - Checked for any incorrect or inconsistent values (e.g. no negative values in Medical insurance charges)
  - Converted categorical features into factor format in R for better interpretability within R

```{r}
colSums(is.na(df))
```
- No missing values in the dataset

# Data

```{r echo = FALSE, tab.cap = "Table 1. Dataset Description"}
kable(data.frame(
  "Feature" = c("smoker", "sex", "region", "charges"),
  "Description" = c(
    "Smoking Status of the individual",
    "Sex of the individual",
    "Region of the individual",
    "Medical insurance charges of the individual"
  ),
  "Data Type and Values" = c(
    "Nominal Categorical, yes/no",
    "Nominal Categorical, female/male",
    "Nominal Categorical, southwest/southeast/northwest/northeast",
    "Discrete Numerical, Positive Integers"
  )
)) %>%
  kable_styling(
    full_width = TRUE, 
    font_size = 16
  )
```

```{r}
summary(df$charges)
```

- Medical insurance charges do not have any negative values, showing no error in the data
- Will dive deeper into their distribution in the later slides

# Descriptive Statistics and Visualisation (Categorical Features)

```{r include = FALSE}
cat_summary_stats <- function(df, cat_feat, color_code) {
  count_df <- as.data.frame(table(df[[cat_feat]]))
  names(count_df) <- c(cat_feat, "Count")
  
  plot <- ggplot(df, aes(x = .data[[cat_feat]], fill = .data[[cat_feat]])) + 
    geom_bar(width = .5) +
    geom_text(
      data = count_df,
      aes(x = .data[[cat_feat]], y = Count, label = Count),
          vjust = -.5, size = 2) +
    labs(title = paste("Distribution of", cat_feat),
         x = cat_feat, y = "Count") +
    scale_fill_manual(values = color_code) +
    theme(axis.text.x = element_blank(), axis.ticks.y = element_blank(), 
          plot.title = element_text(hjust = .5))
  
  return(plot)
}
```

```{r echo = FALSE, fig.cap = "Fig 1. Distribution for Categorical Features", fig.width = 9, fig.height = 3}
(cat_summary_stats(df, "smoker", smoker_color_code) +
  plot_spacer() +
  cat_summary_stats(df, "sex", sex_color_code) +
  plot_spacer() +
  cat_summary_stats(df, "region", region_color_code)) +
  plot_layout(widths = c(1, .05, 1, .05, 1))
```

- Distribution of sex and region and quite balanced across the different categories
- More non-smokers than smokers within the dataset
- Imbalance may reflect a more realistic population behaviour, where smoking is less common due to its known negative health impacts

# Descriptive Statistics and Visualisation (Numerical Features)

```{r echo = FALSE, fig.cap = "Fig 2. Distribution of Medical Insurance Charges", fig.width = 8, fig.height = 2}
boxplot_charges <- ggplot(df, aes(x = charges, y = "")) + 
  geom_boxplot(width = .25, fill = "firebrick1", alpha = .5) +
  labs(x = "charges", y = "") +
  theme(axis.ticks.y = element_blank(), 
        plot.title = element_text(hjust = .5))

histogram_charges <- ggplot(df, aes(x = charges)) +
  geom_histogram(alpha = .5, fill = "firebrick1", position = "identity", bins = 50) +
  labs(x = "Charges", y = "Count", fill = "Smoker") +
  theme(axis.ticks.y = element_blank(), 
        plot.title = element_text(hjust = .5))

(boxplot_charges + plot_spacer() + histogram_charges) +
  plot_annotation(
    title = "Distribution of Medical Insurance Charges",
    theme = theme(plot.title = element_text(hjust = .5))
  ) +
  plot_layout(widths = c(1, .05, 1))
```

```{r echo = FALSE, tab.cap = "Table 2. Descriptive Statistics for Medical Insurance Charges"}
num_des_stat_tab <- df %>% summarise(
  Minimum = min(charges, na.rm = TRUE),
  Q1 = quantile(charges, .25, na.rm = TRUE),
  Median = median(charges, na.rm = TRUE),
  Q3 = quantile(charges, .75, na.rm = TRUE),
  Max = max(charges, na.rm = TRUE),
  IQR = IQR(charges, na.rm = TRUE),
  Mean = mean(charges, na.rm = TRUE),
  SD = sd(charges, na.rm = TRUE)
)

kable(num_des_stat_tab) %>%
  kable_styling(
    full_width = TRUE, 
    font_size = 16
  )
```

- Distribution is very right skewed, with a small percentage of individuals having very high medical insurance charges
- Majority of individuals have medical insurance charges around 4740 to 16639
- The standard deviation is larger at 12110 than IQR 11899, meaning that the variability is also affected by high medical insurance charges

# Categorical association (between Region and Sex)

```{r include = FALSE}
boxplot_comparsion <- function(df, x, y, color_code) {
  count_df <- as.data.frame(table(df[[x]], df[[y]]))
  names(count_df) <- c(x, y, "Count")
  
  plot <- ggplot(count_df, 
                 aes(x = .data[[x]], y = Count, 
                     fill = .data[[y]], group = .data[[y]])) +
    geom_bar(stat = "identity", position = "dodge") +
    geom_text(aes(label = Count),
              position = position_dodge(width = .8), vjust = -.5, size = 2) +
    labs(title = paste("Proportion of", y, "by", x),
         x = x, y = "Count", fill = y) +
    scale_fill_manual(values = color_code) +
    theme(axis.ticks.y = element_blank(), 
          plot.title = element_text(hjust = .5))
  
  print(plot)
}
```

```{r include = FALSE}
chi_square_test <- function(df, feature_1, feature_2) {
  chi2 <- chisq.test(table(df[[feature_1]], df[[feature_2]]))
  
  obs <- chi2$observed
  exp <- round(chi2$expected, 3)
  
  chi2_obs_exp_table <- as.data.frame(
    matrix(
      paste(obs, " (", exp, ")", sep = ""),
      nrow = nrow(obs),
      ncol = ncol(obs),
      dimnames = list(
        rownames(obs),
        colnames(obs)
      )
    )
  )
  
  chi2_obs_exp_table <- cbind(
    setNames(data.frame(rownames(chi2_obs_exp_table)), feature_1),
    chi2_obs_exp_table
  )
  
  rownames(chi2_obs_exp_table) <- NULL
  
  list(
    table = chi2_obs_exp_table,
    results = chi2
  )
}
```

- Chi-square test of independence analysis to examine whether there is an association between sex and region, if the distribution of males and females differ across the different regions
- Identify if the demographic structure is consistent across regions, testing for biases
- H0: Sex and region are independent (No association)
- HA: Sex and region are associated
- Significance level: alpha = 0.05

```{r echo = FALSE, fig.cap = "Fig 3. Proportion of Sex by Region", fig.width = 4, fig.height = 3}
boxplot_comparsion(df, "region", "sex", sex_color_code)
```

- Visually, it does not look like there is a difference in sex across regions

# Categorical association (between Region and Sex)

```{r include = FALSE}
chi2_region_sex <- chi_square_test(df, "region", "sex")
```

```{r echo = FALSE, tab.cap = "Table 3. Contigency Table of Sex by Region. Observed vs (Expected)"}
kable(chi2_region_sex$table) %>%
  kable_styling(
    full_width = TRUE, 
    font_size = 16
  )
```

```{r echo = FALSE}
chi2_region_sex$results
```

- Contingency table:
  - Met assumptions where no more than 25% of cells have expected counts < 5
  - showed observed frequencies close to expected frequencies (in brackets)
- Chi-square test showed p-value = 0.93, which is greater than signifance level at 0.05.
- Fail to reject null hypothesis, indicating there is no statisitcally significant association between sex and region

# Categorical association (between Smoker Status and Sex)

- Chi-square test of independence analysis to examine whether there is an association between sex and smoking status, if the distribution of males and females differ across the different smoking statuses
- Determine if smoking behaviour differs between males and females
- H0: Sex and smoking status are independent (No association)
- HA: Sex and smoking status are associated
- Significance level: alpha = 0.05

```{r echo = FALSE, fig.cap = "Fig 4. Proportion of Sex by Smoking Status", fig.width = 4, fig.height = 3}
boxplot_comparsion(df, "smoker", "sex", sex_color_code)
```

- Visible difference, where males are more likely to smoke than females

# Categorical association (between Smoker Status and Sex)

```{r include = FALSE}
chi2_smoker_sex <- chi_square_test(df, "smoker", "sex")
```

```{r echo = FALSE, tab.cap = "Table 4. Contigency Table of Sex by Smoking Status Observed vs (Expected)"}
kable(chi2_smoker_sex$table) %>%
  kable_styling(
    full_width = TRUE, 
    font_size = 16
  )
```

```{r echo = FALSE}
chi2_smoker_sex$results
```

- Contingency table:
  - Met assumptions where no more than 25% of cells have expected counts < 5
  - showed observed frequencies are slightly further apart to expected frequencies (in brackets)
  - Females were observed to smoke less than expected
  - Male were observed to smoke more than expected
- Chi-square test showed p-value = 0.006, which is less than significance level at 0.05. Reject null hypothesis, indicating there is a statisitcally significant association between sex and smoking status. Males are more likely to smoke than Females

# Hypothesis Testing (between Medical Insurance Charges and Smoking)

```{r include = FALSE}
cat_charges_graphs <- function(df, cat_feat, color_code, cat_title) {
  boxplot_plot <- ggplot(df, aes(x = charges, y = .data[[cat_feat]], 
                                 fill = .data[[cat_feat]])) + 
    geom_boxplot(width = .25) +
    scale_fill_manual(values = color_code) +
    labs(x = "Charges", y = cat_feat, fill = cat_feat) +
    theme(legend.position = "none")
  
  hist_plot <- ggplot(df, aes(x = charges, fill = .data[[cat_feat]])) +
    geom_histogram(alpha = .5, position = "identity", bins = 50) +
    scale_fill_manual(values = color_code) +
    labs(x = "Charges", y = "Count", fill = cat_feat)
  
  (boxplot_plot + plot_spacer() + hist_plot) +
    plot_annotation(
      title = paste("Medical insurance charges by", cat_title),
      theme = theme(plot.title = element_text(hjust = .5))
    ) +
    plot_layout(widths = c(1, .05, 1))
}
```

- Independent two-sample T-test analysis if medical insurance charges differs between smoking status
- Identify if smoking is associated with higher medical insurance costs
- H0: Mean medical insurance charges are equal for all smoking groups
- HA: Mean medical insurance charges differ between smokers and non-smokers
- Significance level: alpha = 0.05

```{r echo = FALSE, fig.cap = "Fig 5. Distribution of Medical Insurance Charges by Smoking Status", fig.width = 6, fig.height = 2}
cat_charges_graphs(df, "smoker", smoker_color_code, "Smoking Status")
```

- Visible difference, where smokers have a higher medical insurance charge

# Hypothesis Testing (between Medical Insurance Charges and Smoking)

```{r include = FALSE}
df_smoke <- df %>% filter(smoker == "yes")
df_non_smoke <- df %>% filter(smoker == "no")
```

```{r include = FALSE}
norm_test <- function(df, num_feat, title) {
  plot <- ggplot(df, aes(sample = .data[[num_feat]])) +
    stat_qq() + stat_qq_line(color = "red") +
    labs(title = title,
         x = "Normal Quantiles", y = "Data Quantiles") +
    theme(plot.title= element_text(hjust = .5))
  
  return(plot)
}
```

```{r echo = FALSE, fig.cap = "Fig 6. Normality Test with QQ Plot for Medical Insurance Charges by Smoking Status", fig.width = 6, fig.height = 2}
(norm_test(df_smoke, "charges", title ="Smokers") + 
    plot_spacer() + 
    norm_test(df_non_smoke, "charges", title = "Non-smokers")) +
  plot_annotation(
    title = "QQ Plot of Medical insurance charges",
    theme = theme(plot.title = element_text(hjust = .5))
  ) +
  plot_layout(widths = c(1, .05, 1))
```

```{r echo = FALSE, tab.cap = "Table 5. Homogeneity of Variance Test for Medical Insurance Charges by Smoking Status"}
kable(leveneTest(charges ~ smoker, data = df), digits = 100) %>%
  kable_styling(
    full_width = TRUE, 
    font_size = 16
  )
```

- Assumptions:
  - Normality test with Q-Q plot showed the data is not normally distributed, however since both smoking groups had sample sizes > 30, T-test still remains valid under Central Limit Theorem
  - Homogeneity of variance test showed uneual variances as the p-value < 0.05
    - Instead of conducting the standard T-test, we will have to conduct the Welch two-sample T-test instead

# Hypothesis Testing (between Medical Insurance Charges and Smoking)

```{r}
t.test(
  charges ~ smoker, data = df,
  var.equal = FALSE, mu = 0, alternative = "two.sided"
)
```

- Welch T-test showed p-value = 2.2e-16, which is less than significance level. Reject null hypothesis, indicating there is a statistically significant difference in mean medical insurance charges between smokers and non-smokers
- Smokers incur significantly higher medical insurance charges than non-smokers, suggesting that smoking might have an association with increased medical costs

# Hypothesis Testing (between Medical Insurance Charges and Sex)

- Independent two-sample T-test analysis if medical insurance charges differs between sex
- Identify if the different sex is associated with higher medical insurance costs
- H0: Mean medical insurance charges are equal for all sexes
- HA: Mean medical insurance charges differ between sexes
- Significance level: alpha = 0.05


```{r echo = FALSE, fig.cap = "Fig 7. Distribution of Medical Insurance Charges by Sex", fig.width = 6, fig.height = 2}
cat_charges_graphs(df, "sex", sex_color_code, "Sex")
```

- Not much visible difference, however the Q1-Q3 range for medical insurance charges is larger for Male than Female

```{r include = FALSE}
df_female <- df %>% filter(sex == "female")
df_male <- df %>% filter(sex == "male")
```

# Hypothesis Testing (between Medical Insurance Charges and Sex)

```{r echo = FALSE, fig.cap = "Fig 8. Normality Test with QQ Plot for Medical Insurance Charges by Sex", fig.width = 6, fig.height = 2}
(norm_test(df_female, "charges", title ="Female") + 
    plot_spacer() + 
    norm_test(df_male, "charges", title = "Male")) +
  plot_annotation(
    title = "QQ Plot of Medical insurance charges",
    theme = theme(plot.title = element_text(hjust = .5))
  ) +
  plot_layout(widths = c(1, .05, 1))
```

```{r echo = FALSE, tab.cap = "Table 6. Homogeneity of Variance Test for Medical Insurance Charges by Sex"}
kable(leveneTest(charges ~ sex, data = df), digits = 100) %>%
  kable_styling(
    full_width = TRUE, 
    font_size = 16
  )
```

- Assumptions:
  - Normality test with Q-Q plot showed the data is not normally distributed, however since both sexes had sample sizes > 30, T-test still remains valid under Central Limit Theorem
  - Homogeneity of variance test showed uneual variances as the p-value < 0.05
    - Instead of conducting the standard T-test, we will have to conduct the Welch two-sample T-test instead
    
# Hypothesis Testing (between Medical Insurance Charges and Sex)

```{r}
t.test(
  charges ~ sex, data = df,
  var.equal = FALSE, mu = 0, alternative = "two.sided"
)
```

- Welch T-test showed p-value = 0.03584, which is less than significance level. Reject null hypothesis, indicating there is a statistically significant difference in mean medical insurance charges between sexes
- Males incur significantly higher medical insurance charges than Females, although this difference might be partly explained through other factors such as the higher proportion of smokers for males.

# Discussion

- Findings:
  - Dataset is unbiased in terms of data collected between sex and region
  - Significant association was found between sex and smoking status, where males are more likely to smoke than females
  - Smokers had significantly higher medical insurance charges than non-smokers
  - While on average that males also had a higher medical insurance charge than females, it can be due to the difference explained between the higher proportion of smokers among males
- Strengths and Limitations:
  - Combined categorical association and hypothesis testing to investigate a couple of perspectives on how factors relate to medical insurance charges
  - Hypothesis stated and assumption checks are always performed before conducting statistical tests
  - Results are also supported with statistical analysis and graphical visualisations
  - However, only selected features were analysed, while other factors (e.g. age and BMI) where not considered
  - Regression analysis not conducted, lose information such as having a combination of multiple features that significantly affect medical insurance charges
  
# Conclusion

- Regardless of region, smoking appears to play a big role in explaning the difference in medical insurance charges, with more smokers being males than females
- Smoking behaviour could be a key factor in higher medical costs, highlighting the importance of how lifestyle choices can influence in medical expenses and insurance outcomes
- Health organisations can explore ways and interventions to help reduce smoking among both males and females, which may help reduce overall medical insurance charges

- Further Investigation:
  - Perform multiple regression analysis to determine if smoking status is still a significant feature affecting charges, or a combination of other features (inclusive/exclusive of smoking status)
  - Investigate more features on the individual's demographic and lifestyle choices

# References

1. Mosap Abdel-Ghany. (2025). Medical Insurance Cost Dataset. Kaggle. https://www.kaggle.com/datasets/mosapabdelghany/medical-insurance-cost-dataset
2. Hansjörg Neth. (2026). D.3 Basic R colors https://bookdown.org/hneth/ds4psy/D.3-apx-colors-basics.html
3. Yihui Xie, J.J. Allaire, Garrett Grolemund. (2026). 4.2 Slidy presentation. https://pkg.yihui.org/rmarkdown-book/slidy-presentation